---
title: "Parasite specificty determines infection of exotic fish species"
author: "Andrew W. Park"
date: "2023-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F,message=F,warning=F,tidy=T)
```

```{r libLoad}
library(tidyverse)
library(magrittr)
library(ape) #plot tree
library(picante) #evol distinct
```


This analysis uses the data from García-Prieto et al. (2022) to study consequences of parasite specificity.

First, we source the `paraPrep` file, which reads in the data, and harmonizes parasites names and taxonomy

```{r sourceParaPrep}
source(knitr::purl("inv.qmd"),quiet=T)
```

Next, we read in a fish phylogenetic tree

```{r readFishPhylo}
library(ape)
tree <- ape::read.tree("actinopt_12k_raxml.tre")

```



```{r filterAdult}
f %<>% mutate(Stage=stringr::str_replace_all(Stage,"Adulto","Adult"))
fA <- f %>% dplyr::filter(Stage=="Adult")

# remove host or parasite taxa not identified to species
fA %<>% dplyr::filter(grepl("sp\\.",Host_species)==F)
fA %<>% dplyr::filter(grepl("sp\\.",Parasite_species)==F)

hostOfAdultParas <- fA %>% dplyr::select(Host_species) %>% distinct() %>% pull()
hostOfAdultParas <- gsub("_"," ",hostOfAdultParas)
```

We create a host-parasite matrix

```{r createHPmatrix}
# matrix of host-(adult)parasite associations
m <- as.data.frame.matrix(table(fA$Parasite_species,fA$Host_species))
```

Currently 56 missing fish (in database but not phylogeny)

```{r eval=F}
#CURRENTLY EVAL FALSE AS FISHBASE PACKAGE HAS DEVELOPED BUG
# probably a fish synonym issue, but for now remove host species not in fish phylo
missingFish <- setdiff(colnames(m),tree$tip.label)
missingFishFormatted <- gsub("_"," ",missingFish)
library(rfishbase)
synCheck <- rfishbase::synonyms(missingFishFormatted,server="fishbase")
synCheck %<>% dplyr::filter(Status %in% c("misapplied name","synonym")) %>% dplyr::select(synonym,Species)
synCheck %<>% dplyr::mutate(across(.cols=everything(),~str_replace_all(.," ","_")))

idx.1 <- which(fA$Host_species==synCheck$synonym[1])#identifies synomym mathcing in tree
idx.2 <- which(fA$Host_species==synCheck$synonym[2])# does not
idx.3 <- which(fA$Host_species==synCheck$synonym[3])# does not

#remove found fish from missing fish
idx.foundFish <- which(missingFish==synCheck$synonym[1])
missingFish <- missingFish[-idx.foundFish]
missingFishFormatted <- gsub("_"," ",missingFish)


fA$Host_species[idx.1] <- synCheck$Species[1]
#idx.2 and idx.3 didn't recover a useful synonym

# recalculate matrix of host-(adult)parasite associations
m <- as.data.frame.matrix(table(fA$Parasite_species,fA$Host_species))

# for now we remove from host-parasite data those fish species that are not found in the phylogenetic tree
fA %<>% dplyr::filter(!Host_species %in% missingFish)
#reCalc fParas now some species removed
fParas <- fA %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()
```

Are the parasites of exotic fish species (that are also parasites of native fish species) a non-random sample?

```{r}

exoParas <- fA %>% dplyr::filter(Exotic_or_native=="Exotic") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()
natParas <- fA %>% dplyr::filter(Exotic_or_native=="Native") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()

onlyExoParas <- setdiff(exoParas,natParas)


fA %<>% dplyr::filter(!Parasite_species%in%onlyExoParas)


natBoth <- tibble(Parasite_species=exoParas,natBoth="both")
natBoth2 <- tibble(Parasite_species=natParas,natBoth="nat")
natBoth %<>% bind_rows(.,natBoth2)

doubleCounted <- natBoth %>% group_by(Parasite_species) %>% summarize(n=n()) %>% dplyr::filter(n>1) %>% pull(Parasite_species)
forTheChop <- NULL
for (i in 1:dim(natBoth)[1]){
  if (natBoth$Parasite_species[i] %in% doubleCounted & natBoth$natBoth[i]=="nat"){
    forTheChop <- c(forTheChop,i)
  }
}

natBoth %<>% slice(-forTheChop)

# need parasite specificity measured only in native species


fAN <- fA %>% dplyr::filter(Exotic_or_native=="Native")
mNat <- as.data.frame.matrix(table(fAN$Parasite_species,fAN$Host_species))

#perform mpd analysis
library(picante)
phydist <- ape::cophenetic.phylo(tree) 

#reduce phydist to host species in fish database
idx <- which(rownames(phydist) %in% fAN$Host_species)
phydist.nat <- phydist[idx,idx]

#reduce HP association matrix to fish in phydist
idx <- which(colnames(mNat) %in% colnames(phydist.nat))
mNat <- mNat[,idx]
idx <- which(rowSums(mNat)!=0)
mNat <- mNat[idx,]
```

```{r eval=FALSE}
zNat <- picante::ses.mpd(mNat,phydist.nat,null.model="independentswap",runs=1000,abundance.weighted=F)
save(zNat,file="get_zNat.Rda")
```

```{r}
load("get_zNat.Rda")
zNat %<>% mutate(Parasite_species=rownames(.))

paraPhylum <- f %>% dplyr::select(Parasite_species,Phylum_parasite) %>% distinct()
paraClass <- f %>% dplyr::select(Parasite_species,Class_parasite) %>% distinct()

zNat %<>% left_join(.,paraPhylum)
zNat %<>% left_join(.,paraClass)

zNat %<>% left_join(.,natBoth)
zNat %<>% drop_na(Phylum_parasite,mpd.obs.z)
zNat %<>% mutate(sg=if_else(mpd.obs.z>(-1.96),"g","s"))

#density plot of specificity
#zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes","Nematoda")) %>% ggplot(.,aes(x=mpd.obs.z,fill=as.factor(natBoth)))+geom_density(alpha=0.3)+facet_wrap(~Phylum_parasite)

plot.zNatA <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes","Nematoda")) %>% ggplot(.,aes(x=mpd.obs.z,fill=as.factor(natBoth)))+geom_density(alpha=0.5,trim=T,col="gray50")+xlab("Phylogenetic specificity")+scale_fill_manual(name="Host type infected",labels=c("Native and exotic","Native only"),values=c("orange","skyblue"))+ylab("Density")+theme_classic()+ggtitle("Adult parasite stages")+facet_wrap(~Phylum_parasite)+xlim(-6,1.5)+ylim(0,0.6)+theme(legend.position="none")



zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes")) %>% ggplot(.,aes(y=mpd.obs.z,x=as.factor(natBoth)))+geom_violin(aes(fill=as.factor(natBoth)))


zPlat1 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes") & natBoth=="nat") %>% dplyr::select(mpd.obs.z) %>% pull()
zPlat2 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes") & natBoth=="both") %>% dplyr::select(mpd.obs.z) %>% pull()

wilcox.test(zPlat1,zPlat2)

zPlat1 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Nematoda") & natBoth=="nat") %>% dplyr::select(mpd.obs.z) %>% pull()
zPlat2 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Nematoda") & natBoth=="both") %>% dplyr::select(mpd.obs.z) %>% pull()

wilcox.test(zPlat1,zPlat2)

zPlat1 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes","Nematoda") & natBoth=="nat") %>% dplyr::select(mpd.obs.z) %>% pull()
zPlat2 <- zNat %>% dplyr::filter(Phylum_parasite %in% c("Platyhelminthes","Nematoda") & natBoth=="both") %>% dplyr::select(mpd.obs.z) %>% pull()

wilcox.test(zPlat1,zPlat2)

#library(brunnermunzel)

#brunnermunzel.test(zPlat1,zPlat2)

#library(nonpar)
#lepage.test(zPlat1,zPlat2)

#library(misty)

```