---
title: "What is the likely source (native/invasive hosts) of parasite species?"
format: html
editor: visual
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo=F,message=F,warning=F,tidy=T)
library(tidyverse)
library(magrittr)
library(ape) #plot tree
library(picante) #evol distinct
```

This analysis uses the data from García-Prieto et al. (2022) to study consequences of parasite specificity.

First, we source the `paraPrep` file, which reads in the data, and harmonizes parasites names and taxonomy

```{r sourceParaPrep}
source(knitr::purl("paraPrep.qmd",quiet=T))
```

Are the parasites exclusively of exotic fish species taxonomically random?

First, make a taxonomic tree

```{r makeTaxaTree}
frm <- ~kingdom/phylum/class/order/family/genus/species

myTaxa %<>% mutate(species=as.factor(species))
myTaxa %<>% mutate(genus=as.factor(genus))
myTaxa %<>% mutate(family=as.factor(family))
myTaxa %<>% mutate(order=as.factor(order))
myTaxa %<>% mutate(class=as.factor(class))
myTaxa %<>% mutate(phylum=as.factor(phylum))
myTaxa %<>% mutate(kingdom=as.factor(kingdom))

tr <- as.phylo(frm, data = myTaxa, collapse=FALSE)
tr$edge.length <- rep(1, nrow(tr$edge))

```


Classify parasite species as infecting only native species, only exotic species, or both

```{r}
exoParas <- f %>% dplyr::filter(Exotic_or_native=="Exotic") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()
natParas <- f %>% dplyr::filter(Exotic_or_native=="Native") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()

onlyExoParas <- setdiff(exoParas,natParas)
bothHostParas <- intersect(exoParas,natParas)
onlyNatParas <- setdiff(natParas,exoParas)
```


