---
title: "What is the likely source (native/invasive hosts) of parasite species?"
format: html
editor: visual
---

```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo=F,message=F,warning=F,tidy=T)
library(tidyverse)
library(magrittr)
library(ape) #plot tree
library(picante) #evol distinct
```

This analysis uses the data from García-Prieto et al. (2022) to study consequences of parasite specificity.

First, we source the `paraPrep` file, which reads in the data, and harmonizes parasites names and taxonomy

```{r sourceParaPrep}
source(knitr::purl("paraPrep.qmd",quiet=T))
```

Are the parasites exclusively of exotic fish species taxonomically random?

First, make a taxonomic tree

```{r makeTaxaTree}
frm <- ~kingdom/phylum/class/order/family/genus/species

myTaxa %<>% mutate(species=as.factor(species))
myTaxa %<>% mutate(genus=as.factor(genus))
myTaxa %<>% mutate(family=as.factor(family))
myTaxa %<>% mutate(order=as.factor(order))
myTaxa %<>% mutate(class=as.factor(class))
myTaxa %<>% mutate(phylum=as.factor(phylum))
myTaxa %<>% mutate(kingdom=as.factor(kingdom))

tr <- as.phylo(frm, data = myTaxa, collapse=FALSE)
tr$edge.length <- rep(1, nrow(tr$edge))

```


Classify parasite species as infecting only native species, only exotic species, or both

```{r}
exoParas <- f %>% dplyr::filter(Exotic_or_native=="Exotic") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()
natParas <- f %>% dplyr::filter(Exotic_or_native=="Native") %>% dplyr::select(Parasite_species) %>% distinct() %>% pull()

onlyExoParas <- setdiff(exoParas,natParas)
bothHostParas <- intersect(exoParas,natParas)
onlyNatParas <- setdiff(natParas,exoParas)

paraSets <- tibble(para=character(0),set=character(0))
onlyExoBuild <- bind_cols(onlyExoParas,"exo")
onlyExoBuild %<>% dplyr::rename("para"="...1")
onlyExoBuild %<>% dplyr::rename("set"="...2")

onlyNatBuild <- bind_cols(onlyNatParas,"nat")
onlyNatBuild %<>% dplyr::rename("para"="...1")
onlyNatBuild %<>% dplyr::rename("set"="...2")

bothBuild <- bind_cols(bothHostParas,"both")
bothBuild %<>% dplyr::rename("para"="...1")
bothBuild %<>% dplyr::rename("set"="...2")

paraSets %<>% bind_rows(onlyExoBuild)
paraSets %<>% bind_rows(onlyNatBuild)
paraSets %<>% bind_rows(bothBuild)

```


```{r plotTree}

tips <- as.data.frame(tr$tip.label)
tips %<>% dplyr::rename(para=`tr$tip.label`)
#hostSet4join <- taxDist %>% mutate(para=str_replace_all(para,"_"," "))
tips %<>% left_join(.,paraSets)

cladeEXObranches<-which.edge(tr,which(tips$set=="exo"))
cladeNATbranches<-which.edge(tr,which(tips$set=="nat"))
cladeBOTHbranches<-which.edge(tr,which(tips$set=="both"))

my.edge.colors<-rep("gray20",length(tr$edge.length))
my.edge.colors[cladeEXObranches]<-"dodgerblue3"
my.edge.colors[cladeNATbranches]<-"gray80"
my.edge.colors[cladeBOTHbranches]<-"brown"

plot(tr,show.tip.label=T,edge.color=my.edge.colors,edge.width=0.5,direction="rightwards",cex=0.3,label.offset=0)#,x.lim=c(0,1200))

```

Test if taxonomic relatedness of exo parasites are different from other (=nat+both) and if metrics of both are different from nat
Ans: exo are different, both are not

```{r}

m <- ape::cophenetic.phylo(tr)

#compare exo to all
comm <- as.data.frame.matrix(table(paraSets$para,paraSets$set))
comm <- t(comm)
z.exo <- picante::ses.mpd(comm,m)

#compare both to nat
paraSetsSub <- paraSets %>% dplyr::filter(set!="exo")
idx <- which(rownames(m) %in% paraSetsSub$para)
mSub <- m[idx,idx]
commSub <- as.data.frame.matrix(table(paraSetsSub$para,paraSetsSub$set))
commSub <- t(commSub)
z.both <- picante::ses.mpd(commSub,mSub)

```

